---
title: Quantitative Methods for Business and Economics
subtitle: Lecture 4 - Logarithms and Related Topics
author: Daniel Sánchez Pazmiño
date: 2026-01-01
theme: berlin
date-format: "YYYY"
fig-width: 8
fig-height: 4.5
format: beamer
fontsize: 10pt
include-in-header:
    - text: |
        \setbeamercolor{background canvas}{bg=white}
        \setbeamertemplate{caption}[numbered]
        \usecolortheme[named=black]{structure}
        \usepackage{tikz}
        \usepackage{pgfplots}
---

# Logarithms

- The logarithm of a number is the power to which a base must be raised to obtain that number.

$$\log_b(x) = y \iff b^y = x$$

*$\iff$ means "if and only if".

- The most common logarithms are base 10 and base $e$ (Euler's number).

- The natural logarithm is the logarithm with base $e$.
    - It is denoted $\ln(x)$.

- Euler's number is approximately 2.71828.
    - It is an irrational number.
    - It was discovered by the Swiss mathematician Leonhard Euler while studying compound interest (percentages).

# Examples

- $\log_{10}(100) = 2$ because $10^2 = 100$.

- $\log_{15}(225) = 2$ because $15^2 = 225$.

- $\log_{e}(e) = 1$ because $e^1 = e$.

- $\ln(e^2) = 2$ because $e^2 = e^2$.

# Properties of Logarithms

**Product rule**: $\log_b(xy) = \log_b(x) + \log_b(y)$

**Quotient rule**: $\log_b\left(\dfrac{x}{y}\right) = \log_b(x) - \log_b(y)$

**Power rule**: $\log_b(x^y) = y \log_b(x)$

**Change of base formula**: $\log_b(x) = \dfrac{\log_a(x)}{\log_a(b)}$

**Logarithm of 1**: $\log_b(1) = 0$

**Exponent rule**: $b^{\log_b(x)} = x$

# Examples of Properties

- $\log_{10}(1000) = \log_{10}(10 \times 100) = \log_{10}(10) + \log_{10}(100) = 1 + 2 = 3$

- $\log_{10}(1000) = \log_{10}(10^3) = 3 \log_{10}(10) = 3$

- $\log_{10}(1000) = \dfrac{\log_{2}(1000)}{\log_{2}(10)} = 3$

# Logarithms are our friends!

- As tricky as they might seem, logarithms are our friends **because of their properties**.

- When dealing with variable exponentials, logarithms can help us simplify the problem.
    - Use the power rule to bring down the exponent.

- Natural logarithms are particularly useful in economics, business, and finance.
    - They are used in growth models because of Euler's number relationship to percent growth.

- Exponential growth shows up in real life in many ways.
    - Compound interest, sales growth, and customer growth can all follow exponential patterns.
    - Logarithms are tools to deal with these phenomena.

# Logarithmic equations

- Logarithmic equations are equations that involve logarithms
    - Include an unknown variable inside the logarithm.

- To solve a logarithmic equation, we need to use the properties of logarithms to simplify the equation.

- Once the equation is simplified, we can solve for the unknown variable using algebraic techniques.

# Example

$$\log_{10}(x) = 2$$

- To solve this equation, we need to remember that $\log_{10}(100) = 2$.

- Therefore, $x = 100$.

# Equations which use the exponent rule

- The reason why logarithms are useful is that they allow us to solve equations that involve exponentials.

- For example: 

$$2^x = 8$$

- To solve this equation, we can take the logarithm of both sides.

$$\log_2(2^x) = \log_2(8)$$

- Using the exponent rule, we get $x = 3$.

# Exponential growth

- Exponential growth is a process that increases at a constant percentage rate over time.

- It is characterized by a constant percentage growth rate.

- Exponential growth is often used to model compound interest, sales growth, and investment balances.

- The formula for exponential growth is $y = a(1 + r)^t$.

# Example

- Suppose \$1,000 is invested at an annual interest rate of 5%.

- The value of the investment after 10 years is given by the formula $y = 1000(1 + 0.05)^{10}$.

- The value of the investment after 10 years is $1000(1.05)^{10} = 1628.89$.

# Example

- In some cases, we might want to know how long it will take for an investment to reach a target value.

- For example, how long will it take for an investment to double if it grows at a rate of 5% per year?

- We can use the formula $y = a(1 + r)^t$ and solve for $t$.

- If the investment doubles, then $2a = a(1 + r)^t$.

- Therefore, $2 = (1 + 0.05)^t$.

- Taking the logarithm of both sides, we get $\log(2) = t \log(1.05)$.

- Therefore, $t = \dfrac{\log(2)}{\log(1.05)} \approx 14.21$ years.

# Example: continuous growth

- In finance, growth is sometimes modeled as *continuous* rather than annual or monthly.

- A simple continuous-growth formula is

$$y = ae^{rt}$$

- Suppose \$1,000 grows continuously at a rate of 5% per year.

- After 10 years, the value is

$$y = 1000e^{0.05 \cdot 10} = 1000e^{0.5} \approx 1648.72$$

- If we want to know when the investment doubles, set $2a = ae^{0.05t}$.

- Then $2 = e^{0.05t}$, so taking natural logarithms gives

$$\ln(2) = 0.05t$$

- Therefore, $t = \dfrac{\ln(2)}{0.05} \approx 13.86$ years.

# Logarithmic scaling

- Logarithmic scales are used when there is a large range of values.
    - They compress the scale to make it easier to read.
    - They often help us observe trends that would be hidden on a regular scale.

- Commonly, the y-axis is in logarithmic scale.
    - Base 10 and natural logarithmic scales are the most common.

# Base 10 logarithmic scale

- The base 10 logarithmic scale is used whenever the data is very large
    - It works especially well when values span powers of 10, i.e., 10, 100, 1000, etc.

- The scale compresses the data to make it easier to read.

- We interpret the values in the scale as powers of 10, since $\log(10^x) = x$.

# Example of base 10 logarithmic scale

- Below, a dataset is shown with values that increase exponentially.
```{r}
library(ggplot2)
library(knitr)

data <- data.frame(x = 1:10, y = 10^seq(1, 10, by = 1))

data$y <- scales::comma(data$y)

kable(data)

```

# Example of base 10 logarithmic scale

- The following plot shows the same data in a base 10 logarithmic scale and a regular scale.

```{r}
#| label: log10-scale

data <- data.frame(x = 1:10, y = 10^seq(1, 10, by = 1))

base_10 <-
  ggplot(data, aes(x = x, y = y)) +
  geom_line() +
  scale_y_log10() +
  labs(y = "Base 10 scale",
       title = "Base 10 logarithmic scale")

natural <-
  ggplot(data, aes(x = x, y = y)) +
  geom_line() +
  labs(y = "Normal scale",
       title = "Normal scale")

gridExtra::grid.arrange(base_10, natural, ncol = 2)

```

# Natural logarithmic scale

- The natural logarithmic scale is often used when the data shows an exponential or percentage growth pattern.

- Powers of $e$ are used to interpret the values in the scale, and these are closely related to continuous percentage growth. 

- If a variable grows exponentially, the trend will not look linear on a regular scale; however, on a natural logarithmic scale, it will look linear.

- Often used in economics, biology, and other fields where exponential growth is common.

# Example of natural logarithmic scale

- Below, a dataset is shown with values that increase exponentially.

```{r}
data <- data.frame(x = 1:10, y = exp(seq(1, 10, by = 1)))

data$y <- scales::comma(data$y)

kable(data)

```

- Notice that the values increase exponentially
    - The percent growth is around 171.8% for each step.

# Example of natural logarithmic scale

- The following plot shows the same data in a natural logarithmic scale and a regular scale.

```{r}

data <- data.frame(x = 1:10, y = exp(seq(1, 10, by = 1)))

natural_log <-
  ggplot(data, aes(x = x, y = y)) +
  geom_line() +
  scale_y_continuous(trans = scales::log_trans(base = exp(1))) +
  labs(y = "Natural logarithmic scale",
       title = "Natural logarithmic scale")

natural <-
    ggplot(data, aes(x = x, y = y)) +
    geom_line() +
    labs(y = "Normal scale",
         title = "Normal scale")

gridExtra::grid.arrange(natural_log, natural, ncol = 2)

```

# $\ln(x)$ is a *very* good friend!

- The assumption of linearity in the social sciences comes up often

- This would mean forcing real-world data such as firm sales or prices to fit a constant increase in dollars each period.

- This is not always realistic, because many business variables grow by percentages rather than by fixed amounts.

- Assuming that, at least over some periods, sales or asset values grow at a constant *percentage* rate is often more realistic.

- Applying a natural logarithm to the data can make the relationship "log-linear", which makes the linearity assumption more realistic.
    - This is one reason $\ln(x)$ appears so often in applied economics and the social sciences.
